### 2018-07-13 submission/manuscript.Rmd
update figure numbers

We next examined whether chimeras also were dependent on polymerase and whether this could affect the total number of OTUs. We observed significant differences in the chimera prevalence based on polymerase across the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S2]. Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S3]. Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 2B & 2C]. The number of cycles used clearly increased chimera prevalence for all polymerases but the rate of increase differed [Figure 2B & 2C]. Additionally, there was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 2D]. A positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 2E]. This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 2E]. This data suggests that chimera prevalence depends on polymerase, is made worse by increasing the number of cycles, and confirms previous reports that the number of OTUs is dependent on the prevalence of these chimeric sequences.

@@

We next investigated the effect polymerases and number of cycles had on the bacterial community composition in more detail by using the Bray-Curtis index. We observed no difference in the bacterial community composition between any measure tested when using the mock community (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 4A]. Finding minimal differences in overall bacterial community composition, we next examined if specific taxa vary depending on polymerase and number of cycles. We found that OTUs that taxonomically classified to *Salmonella*, *Escherichia*, *Enterococcus*, and *Staphylococcus* differed based on polymerase used [Table S6 & S7]. These specific OTUs were all members of the mock community and the differences were most pronounced at 25 and 30 cycles [Table S7]. The majority of the differences between taxa were due to differences between a single polymerase (Kappa) and the other polymerases tested [Table S7].

@@

***Few differences based on number of cycles and polymerase used were identified in fecal samples.***
Unlike mock community samples, an overall difference in the number of OTUs was only detected between polymerases at 35 cycles in fecal samples (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S4], and a Tukey post-hoc test failed to identify any differences between specific polymerases used (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S5]. However, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g. Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 3B]. There was also a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases (e.g. Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 3B]. Based on these observations, there are only small differences in the total number of OTUs detected in fecal samples that are based on number of cycles or polymerases used.

@@

Finally, we built Random Forest models to classify samples for the number of cycles, polymerase used, or individual it was obtained from. Using 16S rRNA gene sequencing data there was a large disparity in the model's ability to correctly classify samples based on polymerase used, number of cycles, and individual (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal") & str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`). There was a clear difference in the success of the model used to classify individuals (probability of correct classification = `r round(exp(-median(fecal_sample_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_sample_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_sample_rf_summary$logLoss)), digits = 2)`)) versus the models created to classify number of cycles or polymerase (cycles model, probability of correct classification = `r round(exp(-median(fecal_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) and polymerase model, probability of correct classification = `r round(exp(-median(fecal_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_poly_rf_summary$logLoss)), digits = 2)`)). Additionally, the top 10 most important OTUs to the polymerase and number of cycle models were not the most important OTUs to the model that classified individuals [Figure 5B]. This suggests that OTUs that change the most due to the number of cycles and polymerase in fecal samples are not the same as OTUs that are part of the biological variation between individuals [Figure 5B]. Collectively, our observations show that despite a variety of errors associated with the amplification process, it is still possible to detect the bacterial community variation between individuals.

>>>>

We next examined whether chimeras also were dependent on polymerase and whether this could affect the total number of OTUs. We observed significant differences in the chimera prevalence based on polymerase across the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S2]. Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S3]. Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 2B & 2C]. The number of cycles used clearly increased chimera prevalence for all polymerases but the rate of increase differed [Figure 2B & 2C]. There was also a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 2D]. Additionally, a positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 2E]. This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 2E]. This data suggests that chimera prevalence depends on polymerase, is made worse by increasing the number of cycles, and confirms previous reports that the number of OTUs is dependent on the prevalence of these chimeric sequences.

@@

We next investigated the effect polymerases and number of cycles had on the bacterial community composition in more detail by using the Bray-Curtis index. We observed no difference in the bacterial community composition between any measure tested when using the mock community (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B]. Finding minimal differences in overall bacterial community composition, we next examined if specific taxa vary depending on polymerase and number of cycles. We found that OTUs that taxonomically classified to *Salmonella*, *Escherichia*, *Enterococcus*, and *Staphylococcus* differed based on polymerase used [Table S6 & S7]. These specific OTUs were all members of the mock community and the differences were most pronounced at 25 and 30 cycles [Table S7]. The majority of the differences between taxa were due to differences between a single polymerase (Kappa) and the other polymerases tested [Table S7].

@@

***Few differences based on number of cycles and polymerase used were identified in fecal samples.***
Unlike mock community samples, an overall difference in the number of OTUs was only detected between polymerases at 35 cycles in fecal samples (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S4], and a Tukey post-hoc test failed to identify any differences between specific polymerases used (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S5]. However, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g. Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 4A]. There was also a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases (e.g. Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 4A]. Based on these observations, there are only small differences in the total number of OTUs detected in fecal samples that are based on number of cycles or polymerases used.

@@

Finally, we built Random Forest models to classify samples for the number of cycles, polymerase used, or individual it was obtained from. Using 16S rRNA gene sequencing data there was a large disparity in the model's ability to correctly classify samples based on polymerase used, number of cycles, and individual (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal") & str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`). There was a clear difference in the success of the model used to classify individuals (probability of correct classification = `r round(exp(-median(fecal_sample_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_sample_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_sample_rf_summary$logLoss)), digits = 2)`)) versus the models created to classify number of cycles or polymerase (cycles model, probability of correct classification = `r round(exp(-median(fecal_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) and polymerase model, probability of correct classification = `r round(exp(-median(fecal_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_poly_rf_summary$logLoss)), digits = 2)`)). Additionally, the top 10 most important OTUs to the polymerase and number of cycle models were not the most important OTUs to the model that classified individuals [Figure 5]. This suggests that OTUs that change the most due to the number of cycles and polymerase in fecal samples are not the same as OTUs that are part of the biological variation between individuals [Figure 5]. Collectively, our observations show that despite a variety of errors associated with the amplification process, it is still possible to detect the bacterial community variation between individuals.

### 2018-07-13 submission/manuscript.tex
update figure numbers

We next examined whether chimeras also were dependent on polymerase and
whether this could affect the total number of OTUs. We observed
significant differences in the chimera prevalence based on polymerase
across the number of cycles used (P-value \textless{} 0.05) {[}Table
S2{]}. Differences in chimera prevalence between Platinum and all other
polymerases accounted for the majority of these differences {[}Table
S3{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
of all polymerases regardless of whether chimera removal with VSEARCH
was used {[}Figure 2B \& 2C{]}. The number of cycles used clearly
increased chimera prevalence for all polymerases but the rate of
increase differed {[}Figure 2B \& 2C{]}. Additionally, there was a
plateau in the total percent of chimeras that were removed that was
similar for all polymerases {[}Figure 2D{]}. A positive correlation was
observed between chimeric sequences and the number of OTUs for all
polymerases {[}Figure 2E{]}. This positive correlation was strongest for
Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 2E{]}.
This data suggests that chimera prevalence depends on polymerase, is
made worse by increasing the number of cycles, and confirms previous
reports that the number of OTUs is dependent on the prevalence of these
chimeric sequences.

@@

We next investigated the effect polymerases and number of cycles had on
the bacterial community composition in more detail by using the
Bray-Curtis index. We observed no difference in the bacterial community
composition between any measure tested when using the mock community
(P-value \textgreater{} 0.05) {[}Figure 4A{]}. Finding minimal
differences in overall bacterial community composition, we next examined
if specific taxa vary depending on polymerase and number of cycles. We
found that OTUs that taxonomically classified to \emph{Salmonella},
\emph{Escherichia}, \emph{Enterococcus}, and \emph{Staphylococcus}
differed based on polymerase used {[}Table S6 \& S7{]}. These specific
OTUs were all members of the mock community and the differences were
most pronounced at 25 and 30 cycles {[}Table S7{]}. The majority of the
differences between taxa were due to differences between a single
polymerase (Kappa) and the other polymerases tested {[}Table S7{]}.

@@

\textbf{\emph{Few differences based on number of cycles and polymerase
used were identified in fecal samples.}} Unlike mock community samples,
an overall difference in the number of OTUs was only detected between
polymerases at 35 cycles in fecal samples (F-stat \textgreater{} 16.35,
P-value = 9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to
identify any differences between specific polymerases used (P-value
\textgreater{} 0.05) {[}Table S5{]}. However, differences in the range
of the number of OTUs detected for fecal samples was dependent on the
polymerase used (e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion
at 35 cycles = 84 - 136) {[}Figure 3B{]}. There was also a trend for
lower number of cycles (15-20) to result in a reduced range in the
number of OTUs versus higher number of cycles (25, 30, and 35) for all
polymerases (e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35
cycles = 84 - 136) {[}Figure 3B{]}. Based on these observations, there
are only small differences in the total number of OTUs detected in fecal
samples that are based on number of cycles or polymerases used.

@@

Finally, we built Random Forest models to classify samples for the
number of cycles, polymerase used, or individual it was obtained from.
Using 16S rRNA gene sequencing data there was a large disparity in the
model's ability to correctly classify samples based on polymerase used,
number of cycles, and individual (P-value \(\leqslant\) 5.49e-07). There
was a clear difference in the success of the model used to classify
individuals (probability of correct classification = 0.87 (0.86 - 0.9))
versus the models created to classify number of cycles or polymerase
(cycles model, probability of correct classification = 0.26 (0.25 -
0.27)(min - max) and polymerase model, probability of correct
classification = 0.18 (0.17 - 0.2)). Additionally, the top 10 most
important OTUs to the polymerase and number of cycle models were not the
most important OTUs to the model that classified individuals {[}Figure
5B{]}. This suggests that OTUs that change the most due to the number of
cycles and polymerase in fecal samples are not the same as OTUs that are
part of the biological variation between individuals {[}Figure 5B{]}.
Collectively, our observations show that despite a variety of errors
associated with the amplification process, it is still possible to
detect the bacterial community variation between individuals.

>>>>

We next examined whether chimeras also were dependent on polymerase and
whether this could affect the total number of OTUs. We observed
significant differences in the chimera prevalence based on polymerase
across the number of cycles used (P-value \textless{} 0.05) {[}Table
S2{]}. Differences in chimera prevalence between Platinum and all other
polymerases accounted for the majority of these differences {[}Table
S3{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
of all polymerases regardless of whether chimera removal with VSEARCH
was used {[}Figure 2B \& 2C{]}. The number of cycles used clearly
increased chimera prevalence for all polymerases but the rate of
increase differed {[}Figure 2B \& 2C{]}. There was also a plateau in the
total percent of chimeras that were removed that was similar for all
polymerases {[}Figure 2D{]}. Additionally, a positive correlation was
observed between chimeric sequences and the number of OTUs for all
polymerases {[}Figure 2E{]}. This positive correlation was strongest for
Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 2E{]}.
This data suggests that chimera prevalence depends on polymerase, is
made worse by increasing the number of cycles, and confirms previous
reports that the number of OTUs is dependent on the prevalence of these
chimeric sequences.

@@

We next investigated the effect polymerases and number of cycles had on
the bacterial community composition in more detail by using the
Bray-Curtis index. We observed no difference in the bacterial community
composition between any measure tested when using the mock community
(P-value \textgreater{} 0.05) {[}Figure 3B{]}. Finding minimal
differences in overall bacterial community composition, we next examined
if specific taxa vary depending on polymerase and number of cycles. We
found that OTUs that taxonomically classified to \emph{Salmonella},
\emph{Escherichia}, \emph{Enterococcus}, and \emph{Staphylococcus}
differed based on polymerase used {[}Table S6 \& S7{]}. These specific
OTUs were all members of the mock community and the differences were
most pronounced at 25 and 30 cycles {[}Table S7{]}. The majority of the
differences between taxa were due to differences between a single
polymerase (Kappa) and the other polymerases tested {[}Table S7{]}.

@@

\textbf{\emph{Few differences based on number of cycles and polymerase
used were identified in fecal samples.}} Unlike mock community samples,
an overall difference in the number of OTUs was only detected between
polymerases at 35 cycles in fecal samples (F-stat \textgreater{} 16.35,
P-value = 9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to
identify any differences between specific polymerases used (P-value
\textgreater{} 0.05) {[}Table S5{]}. However, differences in the range
of the number of OTUs detected for fecal samples was dependent on the
polymerase used (e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion
at 35 cycles = 84 - 136) {[}Figure 4A{]}. There was also a trend for
lower number of cycles (15-20) to result in a reduced range in the
number of OTUs versus higher number of cycles (25, 30, and 35) for all
polymerases (e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35
cycles = 84 - 136) {[}Figure 4A{]}. Based on these observations, there
are only small differences in the total number of OTUs detected in fecal
samples that are based on number of cycles or polymerases used.

@@

Finally, we built Random Forest models to classify samples for the
number of cycles, polymerase used, or individual it was obtained from.
Using 16S rRNA gene sequencing data there was a large disparity in the
model's ability to correctly classify samples based on polymerase used,
number of cycles, and individual (P-value \(\leqslant\) 5.49e-07). There
was a clear difference in the success of the model used to classify
individuals (probability of correct classification = 0.87 (0.86 - 0.9))
versus the models created to classify number of cycles or polymerase
(cycles model, probability of correct classification = 0.26 (0.25 -
0.27)(min - max) and polymerase model, probability of correct
classification = 0.18 (0.17 - 0.2)). Additionally, the top 10 most
important OTUs to the polymerase and number of cycle models were not the
most important OTUs to the model that classified individuals {[}Figure
5{]}. This suggests that OTUs that change the most due to the number of
cycles and polymerase in fecal samples are not the same as OTUs that are
part of the biological variation between individuals {[}Figure 5{]}.
Collectively, our observations show that despite a variety of errors
associated with the amplification process, it is still possible to
detect the bacterial community variation between individuals.
